### 2015-03-14 intro.tex
updates to intro

Device drivers exist at the thin boundary between hardware and software. They are complex pieces of system-level software responsible for the interaction between an operating system and any hardware devices that are attached to a computer. Device drivers are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}. This has a negative impact in hardware product releases, as time-to-market is commonly dominated by device driver development, verification and validation schedules~\cite{yavatkar2012era}.

Even after a device driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

A major reason behind these concurrency bugs is that drivers, as well as other kernel modules in general, operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel is able to invoke its modules from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation. All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.

>>>>

Linux kernel modules (e.g. device drivers and filesystems) exist at the thin boundary between hardware and software. They are complex pieces of system-level software responsible for the interaction between a user application, the operating system and any hardware devices that are attached to a computer. Kernel modules are notoriously hard to develop and even harder to debug~\cite{corbet2005linux}. Especially in the case of drivers, this has a negative impact in hardware product releases, as time-to-market is commonly dominated by device driver development, verification and validation schedules~\cite{yavatkar2012era}.

Even after a driver has shipped, it is typically prone to many serious errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel; while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

A major reason behind these concurrency bugs is that Linux kernel modules operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel is able to invoke its modules from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation. All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.
